Stochastic Approximation Monte Carlo with Dynamic Update Factor (SAD)

### 2018-09-28 papers/histogram/sad.tex
edit paper with Jordan

\subsubsection{Convergence}
The difficulty in using the SAMC method lies in identifying an appropriate
value for $t_0$.  A value that is either too high or too low will result
in slow convergence to the true entropy.  It is instructive to consider
separately values of $t_0$ that are too low or too high.

@@

While for SAMC the update factor is defined in the original
implementation, for SAD the update factor $\gamma_{t}^{\text{SAD}}$ is
thought of as $\nicefrac{\text{d}S}{\text{d}t}$. This tells us that
the SAMC parameter
$t_0$ should have dimensions of entropy.
We begin with an estimate of the average value of the entropy (relative
to the lowest entropy at $T_{\min}$).  If we assume a quadratic
dependence on energy, this is given by
\begin{align}
\langle S\rangle \approx \frac13 \frac{E({T=\infty}) - E(T_{\min})}{T_{\min}}
\end{align}
We approximate this energy difference by
$E_H -E_L$ where $E_H$ and $E_L$ are defined below.
The entropy numerator of the update factor
in general should scale with the total number of interesting
energy states $N_S$, since updates to the weights are distributed between
that many energy states.  The product $N_S\langle S\rangle$ is the
total change of entropy required (starting from constant weights) to find
the true entropy, and puts a lower bound on the convergence time.
After \emph{long} times, when
all the energies have been long ago found,
we wish for a lower $t_0$ in order to more rapidly refine the remaining
error in entropy.  We enable this
by tracking the last time we encountered a new energy $t_L$, and
gradually transition to lower $t_0$ (but still asymptotically scaling
as $\gamma_t \propto 1/t$ to ensure eventual convergence).  Finally,
we wish for an update factor that is \emph{never} greater than 1, because
a very large update factor could introduce errors in entropy that may take
many iterations to remove.  The SAD
expression for $\gamma_t$ which incorporates these ideas is:
\begin{align}
  \gamma_{t}^{\text{SAD}} =
     \frac{
       \frac{E_{H}-E_{L}}{3T_{\text{min}}} + \frac{t}{t_L}
     }{
       \frac{E_{H}-E_{L}}{3T_{\text{min}}} + \frac{t}{N_S}\frac{t}{t_L}
     }
\end{align}
This factor asymptotically has the same $1/t$ behavior as the original
SAMC algorithm;
however now every time a new energy is found to be important
during the course of
the simulation, the update factor will experience a jump. This behavior
allows SAD to dynamically prevent the update factor from decreasing too
rapidly.

@@

\subsection{Convergence}
Although SAD determines the update factor ``dynamically'', the method must still
satisfy the criteria for convergence. Similar to the discussion that was presented
on the convergence for SAMC, we approximate the total entropy change using an integral.
The update factor is integrated from the last time an energy has been discovered $t_L$ to
some minimum time $t_{\min}$ necessary for convergence.
\begin{align}
\Delta S_{\text{tot}} &= \int_{t_L}^{t_{\min}} \gamma dt \\
&\approx N_S \ln\left(\frac{t_{\min}}{t_L}\right)
   + \Delta S_{\text{tot}} \left(1-\frac{t_L}{t_{\min}}\right)
\end{align}
Simplifying, we see that as $t_{\min} \rightarrow \infty$ then $\gamma \rightarrow \infty$ satisfying
the requirement for SAMC convergence. We can solve for the minimum time
\begin{align}
\Delta S_{\text{tot}} \approx N_S \ln\left(\frac{t_{\min}}{t_L}\right)
+ \Delta S_{\text{tot}} \left(1-\frac{t_L}{t_{\min}}\right)
\end{align}
Solving for the minimum time necessary $t_{\min}$, we find that
\begin{align}
\frac{t_{\min}}{t_L} = e^{\frac{\langle S\rangle t_L}{t_{\min}}}
\end{align}
The solution for $t_{\min}$ can be written in terms of the Lambert-W function
\begin{align}
t_{\min} = \frac{3\langle S\rangle}{W(3\langle S\rangle e^{2\langle S\rangle})} t_L
\end{align}
For $\langle S\rangle \rightarrow 0$, we find that the minimum time
$t_{\min} \rightarrow t_L$ and for $\langle S\rangle \rightarrow \infty$,
we find that the minimum time $t_{\min} \rightarrow \frac{3}{2} t_L$.

>>>>

\subsection{SAMC Convergence}\label{sec:samc-convergence}
The difficulty in using the SAMC method lies in identifying an appropriate
value for $t_0$.  A value that is either too high or too low will result
in slow convergence to the true entropy.  It is instructive to consider
separately values of $t_0$ that are too low or too high.

@@

While for SAMC the update factor is defined in the original
implementation, for SAD the update factor $\gamma_{t}^{\text{SAD}}$ is
thought of as $\nicefrac{\text{d}S}{\text{d}t}$. This tells us that
the SAMC parameter
$t_0$ should have dimensions of entropy.
We begin with an estimate of the average value of the entropy (relative
to the lowest entropy at $T_{\min}$).  If we assume a quadratic
dependence on energy, this is given by
\begin{align}
\langle S\rangle \approx \frac13 \frac{E({T=\infty}) - E(T_{\min})}{T_{\min}}
\end{align}
We approximate this energy difference by
$E_H -E_L$ where $E_H$ and $E_L$ are defined below.
The entropy numerator of the update factor
in general should scale with the total number of interesting
energy states $N_S$, since updates to the weights are distributed between
that many energy states.  The product $N_S\langle S\rangle$ is the
total change of entropy required (starting from constant weights) to find
the true entropy, and puts a lower bound on the convergence time.
After \emph{long} times, when
all the energies have been long ago found,
we wish for a lower $t_0$ in order to more rapidly refine the remaining
error in entropy.  We enable this
by tracking the last time we encountered a new energy $t_L$, and
gradually transition to lower $t_0$ (but still asymptotically scaling
as $\gamma_t \propto 1/t$ to ensure eventual convergence).  Finally,
we wish for an update factor that is \emph{never} greater than 1, because
a very large update factor could introduce errors in entropy that may take
many iterations to remove.  The SAD
expression for $\gamma_t$ which incorporates these ideas is:
\begin{align}
  \gamma_{t}^{\text{SAD}} =
     \frac{
       \frac{E_{H}-E_{L}}{T_{\text{min}}} + \frac{t}{t_L}
     }{
       \frac{E_{H}-E_{L}}{T_{\text{min}}} + \frac{t}{N_S}\frac{t}{t_L}
     }
\end{align}
This factor asymptotically has the same $1/t$ behavior as the original
SAMC algorithm;
however now every time a new energy is found to be important
during the course of
the simulation, the update factor will experience a jump. This behavior
allows SAD to dynamically prevent the update factor from decreasing too
rapidly.

@@

\subsection{Convergence}
Although SAD determines the update factor ``dynamically'', the method
must still satisfy the SAMC convergence criteria. As in
Section~\ref{sec:samc-convergence} on the convergence for SAMC, we
will approximate the total entropy change using an integral.  We
integrate update factor from the last time an energy has been
discovered $t_L$ to some minimum time $t_{\min}$ necessary for
convergence.  This gives us a lower bound on the total entropy change
that can be acheived in $t_{\min}$ moves.
\begin{align}
\Delta S_{\text{tot}} &= \int_{t_L}^{t_{\min}}
     \frac{
       3\langle S\rangle + \frac{t}{t_L}
     }{
       3\langle S\rangle + \frac{t}{N_S}\frac{t}{t_L}
     } dt \\
&\approx N_S \ln\left(\frac{t_{\min}}{t_L}\right)
   + \Delta S_{\text{tot}} \left(1-\frac{t_L}{t_{\min}}\right)
\end{align}
where we assumed that $t_L\gg N_S\langle S\rangle$, which is to say
that it takes a long time to find all the interesting energy states.
Simplifying, we see that as $t_{\min} \rightarrow \infty$ then $\gamma
\rightarrow \infty$ satisfying the requirement for SAMC
convergence. We can solve for the minimum time
\begin{align}
\Delta S_{\text{tot}} \approx N_S \ln\left(\frac{t_{\min}}{t_L}\right)
+ \Delta S_{\text{tot}} \left(1-\frac{t_L}{t_{\min}}\right)
\end{align}
Solving for the minimum time necessary $t_{\min}$, we find that
\begin{align}
\frac{t_{\min}}{t_L} = e^{\frac{\langle S\rangle t_L}{t_{\min}}}
\end{align}
The solution for $t_{\min}$ can be written in terms of the Lambert-W function
\begin{align}
t_{\min} = \frac{3\langle S\rangle}{W(3\langle S\rangle e^{2\langle S\rangle})} t_L
\end{align}
For $\langle S\rangle \rightarrow 0$, we find that the minimum time
$t_{\min} \rightarrow t_L$ and for $\langle S\rangle \rightarrow \infty$,
we find that the minimum time $t_{\min} \rightarrow \frac{3}{2} t_L$.
